{\small} High Performance P$^{3}$M N-body code: CUBEP3M:

### 2012-05-23 Extendedpp.tex
Incorporated Ilian annotations on the pdf

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
    \label{fig:power}
  \end{center}
\end{figure}

>>>>

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
  These realizations are evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
 and with $128^{3}$ particles, until $z=0$. The only difference is the ranges of the pp calculation.}
    \label{fig:power}
  \end{center}
\end{figure}

### 2012-05-23 Initial.tex
Incorporated Ilian annotations on the pdf

As mention in section \ref{sec:structure}, the code start off by reading a set of initial conditions.
These correspond to a  $6 \times N$ phase-space array, where $N$ is the number of particles in the
local node. Although most applications so far were based on initial conditions that follow Gaussian statistics,
we have developed a non-Gaussian initial condition generator that we briefly describe in this section.

The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
is extended to include non-Gaussian features of the ``local'' form,
$\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
\phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
to the Bardeen potential $\Phi({\bf x})$ (see \cite{2004PhR...402..103B} for a review). 
We adopted the CMB convention,
in which $\Phi$ is calculated immediately after the matter-
radiation equality (and not at redshift $z=0$ as in the large scale
structure convention). For consistency, $\phi({\bf x})$ is normalized
to the amplitude of scalar perturbations inferred by CMB measurements
($A_s\approx 2.2 \times 10^{-9}$). The local transformation is performed 
before the inclusion of the matter transfer function, and the initial 
particle positions and velocities are finally computed from $\Phi({\bf x})$ 
according to the Zel'dovich approximation, as in the original Gaussian initial condition generator.

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the power spectrum of two realizations with the same initial power spectrum, 
one of which was started of with non-Gaussian initial conditions.
We see that the power spectra agree at the sub-per cent level, as expected {\bf (?????)}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50.eps}
  \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The two curves represent two realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator. The to curves differ at the sub-per cent level, as seen in the top panel.   
    \label{fig:init}}
\end{center}
\end{figure}

>>>>

As mention in section \ref{sec:structure}, the code starts off by reading a set of initial conditions.
These are organized as a set of $6 \times N$ phase-space arrays -- one per {\small MPI} process -- where $N$ is the number of particles in the
local volume. Although many applications are typically based on initial conditions that follow Gaussian statistics,
we have developed a non-Gaussian initial conditions generator that we briefly describe in this section.

The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
is extended to include non-Gaussian features of the `local' form,
$\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
\phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
to the Bardeen potential $\Phi({\bf x})$ (see \cite{2004PhR...402..103B} for a review). 
We adopted the CMB convention,
in which $\Phi$ is calculated immediately after the matter-
radiation equality (and not at redshift $z=0$ as in the large scale
structure convention). For consistency, $\phi({\bf x})$ is normalized
to the amplitude of scalar perturbations inferred by CMB measurements
($A_s\approx 2.2 \times 10^{-9}$). The local transformation is performed 
before the inclusion of the matter transfer function, and the initial 
particle positions and velocities are finally computed from $\Phi({\bf x})$ 
according to the Zel'dovich approximation, as in the original Gaussian initial condition generator.

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the late time power spectrum of two realizations that started off the same initial power spectrum, 
but one of which had non-Gaussian initial conditions.
We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
agrees much better with one loop perturbation theory \citep{XXX}.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50_w_1loop_PT.eps}
  \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The two curves represent two realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator. The to curves differ at the sub-per cent level, as seen in the top panel.   
    \label{fig:init}}
\end{center}
\end{figure}

### 2012-05-23 Introduction.tex
Incorporated Ilian annotations on the pdf

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
 (\cite{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H}; Harnois-D\'{eraps} et al 2012 (in preparation)), 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F},  
and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

@@

The paper is structured as follow: XXXX

>>>>

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
 (\cite{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H}; Harnois-D\'{eraps} et al 2012 ({\bf update with last paper!})), 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F},  
and it is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

%\subsection{Simulation suites}
%\label{subsec:suites}

@@

Tests on constraints of the redshift step size were performed 
on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
Each node contains 2 quad core Intel(R) Xeon(R) E5310 @ 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
We generate a set of Gaussian initial conditions at a starting redshift of $z = 100$, 
with a box size equal to $200 h^{-1}\mbox{Mpc}$, we placed
$128^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
Tests on improvements of the force calculations were performed on the same machine, but with 8 times more particles and node, 
and with a box size of $500 h^{-1}\mbox{Mpc}$.
Hereafter we refer to these simulation modes as the CITA128 and CITA256 configurations respectively.

We then test the dynamical range of the simulation, the non-Gaussian initial conditions generator and the run time halo finder algorithm 
with a third simulation configuration, which we denote the  RANGER4000, that also evolves $4000^{3}$ particles 
from $z=100$ to $z=0$ but with a box side of $3.2 h^{-1}\mbox{Gpc}$.
 Much larger runs were produced at various high performance computing centers
 around the world; their description and a comparison of their performance are also described in this paper.

The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
we then present in section \ref{sec:scaling} the scaling of the code to very large problems;
sections \ref{sec:accuracy} and \ref{sec:systematics} discuss the accuracy and systematic effects of the code respectively.
 We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions to 
 to the default configuration, and conclude afterwards.

### 2012-05-23 PoissonSolver.tex
Incorporated Ilian annotations on the pdf

This section reviews how Poisson's equation is solved on a double-mesh configuration. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
to section 2 of MPT for more details. In {\small CUBEP3M}, the mass default assignment scheme are
a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
for pairs closer to a tenth of a grid cell.

The code units inherit from \citep{2004NewA....9..443T}, which we summarized here for completeness.
In simulation units, the mean comoving mass density is set to unity, the length of a fine grid cell is set to one,
and the Newton gravitational  constant $G$ is set to $1/(6 \pi a)$, where $a$ is the scale factor.
With these three choices, one can transform any physical units into simulation units and vice versa.

>>>>

This section reviews how Poisson's equation is solved on a double-mesh configuration. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
to section 2 of MPT for more details. In {\small CUBEP3M}, the mass default assignment scheme are
a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, 
but we instead impose a sharp cutoff on the force
for pairs that air closer than a tenth of a grid cell.

The code units inherit from \citep{2004NewA....9..443T}, which we summarize here for completeness.
In simulation units,  the comoving length of a fine grid cell is set to one,
such that the unit length in simulation unit is 
\begin{eqnarray}
1\mathcal{L} = a \frac{L}{N} 
\end{eqnarray}
where $a$ is the scale factor, $N$ is the total number of cells along one dimension,
and $L$ is the comoving volume in $h^{-1}\mbox{Mpc}$.
The mean comoving mass density is also set to unity in simulation units, 
which, in physical units, corresponds to 
\begin{eqnarray}
1\mathcal{D} = \rho_{m}(0) a^{-3} = \Omega_{m} \rho_{c} a^{-3} = \frac{3 \Omega_{m} H_{o}^{2}}{8 \pi G a^{3} }
\end{eqnarray}
$\Omega_{m}$ is the matter density, $H_{o}$ is the Hubble's constant, $\rho_{c}$ is the critical density today,
and $G$ is Newton's constant. The mass unit is found with $\mathcal{M} = \mathcal{DL}^{3}$.
Specifying the value of $G$ on the grid fixes the time unit, and with the choice $G_{grid}$ = 1/$(6 \pi a)$,
we get:
\begin{eqnarray}
1 \mathcal{T} = \frac{2a^{2}}{3}\frac{1}{\sqrt{\Omega_{m}H_{o}^{2}}}
\end{eqnarray}
These choices completely determine the convertion between physical units into simulation units and vice versa.
For instance, the velocity units are given by $1\mathcal{V} = \mathcal{L}$/$\mathcal{T}$.

### 2012-05-23 ScalingPerformance.tex
Incorporated Ilian annotations on the pdf

\section{Scaling Performances}

Real time versus size, bottle neck, description of the largest runs, etc.
{\bf (Ilian, I will need your input here as well. Hugh, do you have anything you would like to share 
concerning the largest runs you did on Blue Gene? )}

>>>>

\section{Scaling Performances}
\label{sec:scaling}

\begin{figure*}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_curie.eps}
    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_new.eps}
%  \vskip -1.2cm 
%  \vskip -0.5cm 
  \caption{Scaling of {\small CUBEP3M} on Curie fat nodes (left) and 
    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
    in each case. Dashed line indicates the ideal weak 
    scaling. The data is listed in Table \ref{summary_scaling_table}.
    \label{scaling}
% \vskip -0.9cm 
}
\end{center}
\end{figure*}

\begin{table*}%[ht]
  \vskip -0.5cm 
  \begin{center}
\caption{Scaling of {\small CUBEP3M} on Curie. Speedup is 
scaled to the smallest run.}
\label{summary_scaling_table}
\begin{tabular}{@{}|llllll|}
\hline
number of cores & speedup & ideal speedup & absolute timing (min) & 
$N_{\rm particles}$& box size ($h^{-1}$Mpc)
\\[2mm]\hline
%\hline
32  &  1.00 & - &3.2 & $256^3$ & 256\\
256  & 7.21 & 8 &3.55 & $512^3$  & 512\\
864  & 25.63 & 27 &4.8 & $864^3$  & 864\\
2048  & 61.87 & 64 &26.48 & $2048^3$ & 2048 \\
\hline
\end{tabular}
\caption{Scaling of  {\small CUBEP3M} for large number of cores 
(on Ranger). Speedup is scaled to the smallest run.}
\label{summary_scaling_table2}
\begin{tabular}{@{}|llllll|}
\hline
number of cores & speedup & ideal speedup & absolute timing (min) & 
$N_{\rm particles}$& box size ($h^{-1}$Mpc)
\\[2mm]\hline
%\hline
864    & 1.00  & -    &258   & $1728^3$  & 6.3\\
4096   & 4.53  & 4.74 &320   & $3072^3$  & 11.4\\
10976  & 9.78  & 12.7 &845   & $5488^3$  & 20\\
21952  & 14.73 & 25.4 &561   & $5488^3$  & 20 \\
\hline
\end{tabular}
\end{center}
  \vskip -0.7cm 
\end{table*}

%the Ranger system at the Texas Supercomputing
%Center (in top 20 in the world) which is a SunBlade x6420 
%with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
% ``Barcelona'' processors and Infiniband networking. It has 
%a total of 62976 computing cores and 125952 GB of total 
%memory. Its nodes consist of 4 Quad Core processors and 32 GB 
%of shared RAM. For efficiency reasons (local memory access) we 
%typically use smaller MPI 'nodes' consisting of one Quad Core 
%processor and 8 GB of RAM.

The parallel algorithm of {\small CUBEP3M} is designed for `weak' 
scaling, i.e. if the number of cores and the problem size 
increase in proportion to each other, then for ideal scaling the 
wall-clock time should remain the same, in contrast to `strong' 
scaling, whereby the same problem solved on more cores should take 
proportionately less wall-clock time. This weak scaling requirement 
is dictated by the problems we are typically investigating (very 
large and computationally-intensive) and our goals, which are to 
address such large problems in the most efficient way, rather than 
for the least wall-clock time. Furthermore, there is no explicit 
load balancing, thus the code is most efficient if the sub-domains
contain roughly equal number of particles, which is true for most
cosmological-size volumes, but not for e.g. simulations of a single
highly-resolved galaxy.

%Since the current Curie is using Intel Nehalem architecture, while 
%the thin Curie nodes will use the newer Westermere Intel architecture, 
%we also show the scaling of our code on the Westermere-based Lonestar 
%computer at the Texas Advanced Computing Centre

The scaling was tested with a dedicated series 
of simulations -- the CURIE simulation suite-- with increasing size and number of cores on the `fat' 
(i.e. large-memory) nodes of the supercomputers Curie at the Tr\`{e}s Grand Centre de Calcul (TGCC) in France. 
Our results are shown in Fig. \ref{scaling} and in 
Table \ref{summary_scaling_table}. For appropriate direct comparison,
all these simulations were performed using the same particle mass 
($M_{\rm particle}=1.07\times10^{11}M_\odot$) and force resolution 
(softening length 50 $h^{-1}$kpc). The box sizes used range from 256 $h^{-1}$Mpc
to 2048 $h^{-1}$Mpc, and the number of particles from $256^3$ to $2048^3$.
Simulations were run on 32 up to 2048 computing cores, also starting from 
redshift $z=100$, and evolving until $z=0$. Our results show excellent scaling, within 
$\sim3$ per cent of the ideal one, for up to 2048 cores.

We have also ran {\small CUBEP3M} on a much larger number of cores, 
from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
particles on Ranger at the Texas Advanced Computing Centre (TACC), a SunBlade 
x6420 system with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
`Barcelona' processors and Infiniband networking,%, a total of 62,976 
%computing cores and 125,952 GB of total memory. 
%Its shared-memory nodes consist of 4 Quad Core processors and 32 GB of RAM.) 
and on JUROPA at the J\"ulich Supercomputing Centre in Germany, an Intel Xeon X5570 
(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.%, a total of 17,664
%computing cores and 52 TB of total memory. Its shared-memory nodes 
%consist of 2 Quad Core processors and 24 GB of RAM). 
Since it is not practical to perform dedicated scaling tests on such large number 
computing cores, we instead use the data extracted from production runs, 
as listed in Table \ref{summary_scaling_table2}. We have found the 
code to scale quite well, within 1.5 per cent of ideal up to 4,096 cores. 
For very large number of cores (10,976) the scaling is slightly worse,
due to increased communication 
costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
and load balancing issues, but still within $\sim20$ per cent of ideal. 
These first three Ranger runs were performed 
with 4 {\small MPI} processes and 4 threads per Ranger node ('4way') (using the 
specially-provided {\it tacc\_affinity} NUMA script to bind the memory 
usage to local sockets, thus ensuring memory affinity. This is important 
because the 32 GB RAM/node on Ranger (and other computers with 
multi-processor nodes) are actually not equal-access and the local 
8 GB memory of each processor has much shorter access time).

Furthermore, due to the very strong clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain well above the 
average number of particles, thereby requiring more memory per {\small MPI} process
in order to run. As a consequence, throughout most of the evolution the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 {\small MPI} processes and 8 threads per Ranger node (2way), which on 
Ranger allows using up to 16 GB of RAM per {\small MPI }process. In order to insure 
local memory affinity a special NUMA control script (which we called {\it 
tacc\_affinity\_2way}) 
was developed for us by TACC staff and was used by us to run more efficiently 
in this mode. However, this not-fully-local memory access does affect the 
performance somewhat, as does the imperfect load balancing in such situations,
as seen by the rightmost point on the scaling plot. The scaling in this case
is 42 per cent below the ideal, although we note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus quite satisfactory and it clearly 
scales very well to extremely large number of cores and can therefore 
do very large problems efficiently, utilizing the next generation 
Petascale systems.

Finally, we note that several special fixes had to be developed by TACC 
and JUROPA staff in order for our largest runs to work properly and to be 
able to use the software libraries required ({\small MPICH} and {\small FFTW}), as those 
exhibited serious problems when applied to calculations of such 
unprecedented size.

{\bf Bottle neck in the largest runs?}

{\bf (Hugh, do you have anything you would like to share 
concerning the largest runs you did on Blue Gene? )}

### 2012-05-23 Structure.tex
Incorporated Ilian annotations on the pdf

As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. 
The code has extensions that departs from this basic scheme, and
we shall come back to these later, but for the moment, we adopt the 
standard configuration. 
The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing. In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles},
and for maximal efficiency, the number of tiles per node should be at least equal to the number of available processors.
Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes, as the information exchange between cores that are part of the same motherboard is generally much faster.
As mentioned in MPT, the computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned. The thickness of this surface must be larger than the cutoff length, and we find that a 24 cells deep buffer
is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
the bulk of the foot-print is concentrated in a handful of fine grid arrays.
In addition, some of these are required for intermediate steps of the calculations only, 
hence it is possible to hide some of the coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in a local domain}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\end{enumerate}

>>>>

As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. 
The code has extensions that departs from this basic scheme, and
we shall come back to these later, but for the moment, we adopt the 
standard configuration. 
The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing. In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles},
and for maximal efficiency, the number of tiles per node should be at least equal to the number of available processors.

Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes in relatively small runs, as the information exchange between cores that are part of the same motherboard is generally much faster.
As one scales to larger problems, however, the formation of dense objects can cause memory problems in such configurations, and increasing 
the number of {\small MPI} processes helps to ensure memory locality.

As mentioned in MPT, the computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned. The thickness of this surface must be larger than the cutoff length, and we find that a 24 cells deep buffer
is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}.
Only when it comes to finding haloes at run time can this buffer create a problem, because a halo located close to the boundary
can have a radius larger than the buffer zone.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
the bulk of the foot-print is concentrated in a handful of fine grid arrays.
In addition, some of these are required for intermediate steps of the calculations only, 
hence it is possible to hide some of the coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\end{enumerate}

### 2012-05-23 Systematics.tex
Incorporated Ilian annotations on the pdf

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the p$^3$m algorithm, compared with the exact $1/r^{2}$ law.
  Particles in the same cell follow the exact curve, then we observe a large scatter at 
  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

>>>>

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
  Particles in the same fine cell follow the exact curve, then we observe a large scatter at 
  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

### 2012-05-23 paper.tex
Incorporated Ilian annotations on the pdf

\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M: }
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}, Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}, 
JD Emberson$^{1,5}$ \thanks{E-mail: emberson@astro.utoronto.ca} and Vincent Desjacques$^{6,7}$ \thanks{E-mail:vincent.desjacques@unige.ch}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Ontario, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
$^{4}$Department of Mathematics and Computer Science, Laurentian University, P3E 2C6, Ontario, Canada\\
$^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
$^{6}$Institute for Theoretical Physics, University of Z\"{u}rich, Z\"{u}rich, CH 8057, Switzerland\\
$^{7}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}

@@

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
a two-mesh gravity solver that was already among the fastest N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical.
Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.
We discuss the structure of the code, its accuracy, and its scaling performance.
In addition, many utilities have been added, including a runtime halofinder,
particle identification tags and non-Gaussian initial conditions generator, and we briefly describe their implementation strategy
and accuracy, when applicable .

>>>>

\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M: }
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}, Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}, \newauthor
JD Emberson$^{1,5}$ \thanks{E-mail: emberson@astro.utoronto.ca} and Vincent Desjacques$^{6,7}$ \thanks{E-mail:vincent.desjacques@unige.ch}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Ontario, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
$^{4}$Department of Mathematics and Computer Science, Laurentian University, P3E 2C6, Ontario, Canada\\
$^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
$^{6}$Institute for Theoretical Physics, University of Z\"{u}rich, Z\"{u}rich, CH 8057, Switzerland\\
$^{7}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}

@@

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
solves gravity on a two-level mesh and that is among the fastest publicly available N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical,
which allows for massive parallelization of the code.
Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.
We discuss the structure of the code, its accuracy, and its scaling performance.
In addition, many utilities have been added, including a runtime halofinder,
particle identification tags and non-Gaussian initial conditions generator, and we briefly describe their implementation strategy
and accuracy, when applicable .
